\section{Experimental Setup}
\label{sec:experiments}

\subsection{Datasets}

\begin{table}[t]
\centering
\caption{Dataset summary.}
\label{tab:datasets}
\begin{tabular}{lcccc}
\toprule
Dataset & Modality & Classes & Train & Test \\
\midrule
ACDC~\cite{bernard2018acdc}    & Cardiac MRI   & 3 (RV, Myo, LV) & -- & -- \\
BUSI~\cite{al2020busi}         & Breast US     & 1 (tumour)      & -- & -- \\
FUGC~\cite{fugc2025}           & Colonoscopy   & 2 (polyp, edge) & -- & -- \\
TN3K~\cite{gong2021tn3k}       & Thyroid US    & 1 (nodule)      & -- & -- \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Implementation Details}

\paragraph{Architecture.}
All experiments use a 2-D UNet with channel progression
$[32, 64, 128, 256, 512]$, plain convolutional blocks, batch normalisation,
and dropout ($p=0.1$).

\paragraph{Optimisation.}
Adam optimiser ($\beta_1=0.9$, $\beta_2=0.999$, weight decay $5\times10^{-4}$),
learning rate $\eta=10^{-3}$ with polynomial decay (exponent 0.9) over
$\tau = 200$ local iterations.
Gradient clipping at norm 10.

\paragraph{FL setup.}
$K=5$ clients, $R=10$ communication rounds, $\tau=200$ local iterations per
client per round.
Client participation fraction $C=1.0$ (all clients selected each round).
Three independent runs (seeds 1337, 1338, 1339); results reported as mean.

\paragraph{FedAL setup.}
Initial labelled set $L_0 = 10$ samples per client;
budget $B = 5$ samples queried per round per client;
entropy-based query strategy.

\paragraph{Augmentation.}
ACDC/TN3K: random rotation 90°, mirroring, affine deformation.
BUSI/FUGC: affine (scale $[0.7,1.4]$, degrees $\pm15°$), Gaussian
noise/blur, brightness/contrast jitter, low-resolution simulation, gamma.

\paragraph{Evaluation.}
Dice Similarity Coefficient (DSC, \%, higher is better) and
95th-percentile Hausdorff Distance (HD$_{95}$, mm, lower is better)
computed on the official validation/test splits.
