\section{Conclusion}
\label{sec:conclusion}

We presented a federated learning framework for privacy-preserving medical
image segmentation that unifies four aggregation strategies and introduces
FedAL, a federated active-learning protocol that couples entropy-based
querying with FL communication rounds.
Experiments across four clinically diverse benchmarks demonstrate that
normalisation-based aggregation (FedNova) consistently mitigates client
drift under high data heterogeneity, while personalised aggregation
(FedPer) benefits single-class ultrasound tasks.
FedAL achieves competitive segmentation performance with a fraction of the
labelling budget, offering a practical path to cost-effective FL deployment.

\paragraph{Limitations.}
All experiments simulate federated clients on a single machine; actual
network latency, communication compression, and system heterogeneity are
not modelled.
Privacy is achieved through data locality alone; differential privacy or
secure aggregation are not incorporated, which may be required by certain
regulatory frameworks.

\paragraph{Future work.}
We plan to (i) integrate Gaussian differential privacy noise into the
aggregation step; (ii) extend FedAL with BADGE~\cite{ash2020badge} and
coreset~\cite{sener2018coreset} querying, and investigate the complementary
evidential uncertainty approach of FEAL~\cite{chen2024feal} for
domain-shifted federated clients; (iii) deploy
the system on real hospital nodes using NVIDIA FLARE~\cite{nvidia2019clara};
and (iv) explore 3-D segmentation tasks (e.g., LA2018, BTCV) in the
federated setting.
